The Unicode Cookbook \linebreak for Linguists

### 2018-06-17 book/chapters/implementation.tex
deals with all paperhiver comments except a few time consuming ones
closes #30

Searching the Web for ``install r and python'' returns numerous tutorials on how to set up your machine for scientific data analysis. Note that there is no single correct setup for a particular computer or operating system. Both Python and R are available for Windows, Mac, and Unix operating systems from the Python and R project websites. Another option is to use a so-called package manager, i.e.\ a software program that allows the user to manage software packages and their dependencies. On Mac, we use Homebrew,\footnote{\url{https://brew.sh/}} a simple-to-install (via the Terminal App) free and open source package management system. Follow the instructions on the Homebrew website and then use Homebrew to install R and Python (as well as other software packages such as Git and Jupyter Notebooks).

@@

To install the Python package \texttt{segments} from the Python Package Index (PyPI) run:

@@

The R implementation is available in the package \texttt{qlcData}, which is 
directly available from the central R repository CRAN (Comprehensive R Archive 
Network). The R software environment itself has to be downloaded from its 
website.\footnote{\url{https://www.r-project.org}} After starting the included 
R program, the \texttt{qlcData} package for dealing with orthography profiles can be 
simply installed as follows:

>>>>

Searching the web for ``install r and python'' returns numerous tutorials on how to set up your machine for scientific data analysis. Note that there is no single correct setup for a particular computer or operating system. Both Python and R are available for Windows, Mac, and Unix operating systems from the Python and R project websites. Another option is to use a so-called package manager, i.e.\ a software program that allows the user to manage software packages and their dependencies. On Mac, we use Homebrew,\footnote{\url{https://brew.sh/}} a simple-to-install (via the Terminal App) free and open source package management system. Follow the instructions on the Homebrew website and then use Homebrew to install R and Python (as well as other software packages such as Git and Jupyter Notebooks).

@@

To install the Python package \texttt{segments} \citep{ForkelMoran2018} from the Python Package Index (PyPI) run:

@@

The R implementation is available in the package \texttt{qlcData} \citep{Cysouw2018}, which is 
directly available from the central R repository CRAN (Comprehensive R Archive 
Network). The R software environment itself has to be downloaded from its 
website.\footnote{\url{https://www.r-project.org}} After starting the included 
R program, the \texttt{qlcData} package for dealing with orthography profiles can be 
simply installed as follows:

### 2018-06-17 book/chapters/orthography_profiles.tex
closes #30

At this point in the course of rapid ongoing developments, we are left with a
situation in which the Unicode Standard offers a highly detailed and flexible
approach to deal computationally with writing systems, but it has unfortunately
not influenced the linguistic practice very much. In many practical situations,
the Unicode Standard is far too complex for the day-to-day practice in
linguistics because it does not offer practical solutions for the down-to-earth
problems of many linguists. In this section, we propose some simple practical
guidelines and methods to improve on this situation.

Our central aims for linguistics, to be approached with a Unicode-based
solution, are: (i) to improve the consistency of the encoding of sources, (ii)
to transparently document knowledge about the writing system (including
transliteration), and (iii) to do all of that in a way that is easy and quick to
manage for many different sources with many different writing systems. The
central concept in our proposal is the \textsc{orthography profile}, a simple
delimited text file, that characterizes and documents a writing system.
We also offer basic implementations in Python and R to assist with the
production of such files, and to apply orthography profiles for consistency
testing, grapheme tokenization and transliteration. Not only can orthography
profiles be helpful in the daily practice of linguistics, they also succinctly
document the orthographic details of a specific source, and, as such, might
fruitfully be published alongside sources (e.g.~in digital archives). Also, in
high-level linguistic analyses in which the graphemic detail is of central
importance (e.g.~phonotactic or comparative-historical studies), orthography
profiles can transparently document the decisions that have been taken in the
interpretation of the orthography in the sources used.

Given these goals, Unicode Locales (see
Chapter~\ref{the-unicode-approach}) might seem like the ideal orthography
profiles. However, there are various practical obstacles preventing the use of
Unicode Locales in the daily linguistic practice, namely: (i) the
XML structure is too verbose to easily and quickly produce or correct manually,
(ii) Unicode Locales are designed for a wide scope of information (like date
formats or names of weekdays) most of which is not applicable for documenting
writing systems, and (iii) most crucially, even if someone made the effort to
produce a technically correct Unicode Locale for a specific source at hand,
then it is well-nigh impossible to deploy the description. This is because a locale
description has to be submitted to and accepted by the Unicode Common Locale
Data Repository. The repository is (rightly so) not interested in descriptions
that only apply to a limited set of sources (e.g.~descriptions for only a single
dictionary).

@@

The crucial difference between our current proposal and traditional
computational approaches to transliteration is the strict separation between
tokenization and transliteration. Most computational approaches to
transliteration are based on finite-state transducers (including the
transliteration as described in the Unicode Locale Data Markup
Language).\footnote{See \url{http://www.unicode.org/reports/tr35/}} Finite-state 
transducers attempt to describe the mapping from input to output string directly 
as a set of rewrite rules. Although such systems are computationally well 
understood, we feel that they are not well-suited for day-to-day linguistic 
practice. First, by forcing a first step of grapheme tokenization, our system 
tries to keep close to the logic of the writing system. Second, by separating 
tokenization from transliteration there is no problem with `feeding' and 
`bleeding' of rules, common with transducers (cf.~Section~\ref{r-implementation}).

@@

In practice, we foresee a workflow in which orthography profiles are iteratively
refined, while at the same time inconsistencies and errors in the data to be
tokenized are corrected. In some more complex use-cases there might even be a
need for multiple different orthography profiles to be applied in sequence (see
Sections~\ref{python-implementations} \& \ref{r-implementation} on various exemplary use-cases). The result of any such
workflow will normally be a cleaned dataset and an explicit description of the
orthographic structure in the form of an orthography profile. Subsequently, the
orthography profiles can be easily distributed in scholarly channels alongside
the cleaned data, for example in supplementary material added to journal papers
or in electronic archives.

@@

\begin{enumerate}
	\def\labelenumi{A\arabic{enumi}.} 
	\item \textsc{A profile is a unicode utf-\scalebox{0.8}{8} encoded text file} 
	   normalized following NFC (or NFD if specified in the metadata)
       that includes the information pertinent to the orthography.\footnote{See 
	   Section~\ref{pitfall-file-formats} in which we suggest to use NFC, 
	   no-BOM and LF line breaks because of the pitfalls they avoid. A keen reviewer notes, however, that specifying 
	   a convention for line endings and BOM is overly strict because most 
	   computing environments (now) transparently handle both alternatives. 
	   For example, using Python a file can be decoded using the encoding 
	   ``utf-8-sig'', which strips away the BOM (if present) and reads 
	   an input full in text mode, so that both line feed variants ``LF'' and 
	   ``CRLF'' will be stripped.}

@@

\begin{enumerate}
	\def\labelenumi{A\arabic{enumi}.} \setcounter{enumi}{4} 
	\item \textsc{A minimal profile consists of a single column} with a header
       called \texttt{Grapheme}, listing each of the different graphemes in a
       separate line. The name of this column is crucial for automatic 
       processing.
	\item \textsc{Optional columns can be used to specify the left and right
       context of the grapheme}, to be designated with the headers \texttt{Left}
       and \texttt{Right} respectively. The same grapheme can occur multiple
       times with different contextual specifications, for example to
       distinguish different pronunciations depending on the context. 
	\item \textsc{The columns \texttt{Grapheme}, \texttt{Left} and \texttt{Right}
       can use regular expression metacharacters.} If regular expressions are
       used, then they must be specified in the metadata file as such, 
	   and all literal usage of the special symbols, like full stops <.>
       or dollar signs <\$> (so-called \textsc{metacharacters}) have to be
       explicitly escaped by adding a backslash before them (i.e.~use
       <\textbackslash.> or <\textbackslash\$>). Note that any specification of
       context automatically expects regular expressions, so it is 
       better to always escape all regular expression metacharacters when used
       literally in the orthography. The following symbols will need to be
       preceded by a backslash: {[} {]} ( ) \{ \} | ~+ * . - ! ? \^{} \$ and the
       backslash \textbackslash~itself. 
	\item \textsc{An optional column can be used to specify classes of graphemes},
       to be identified by the header \texttt{Class}. For example, this column
       can be used to define a class of vowels. Users can simply add ad-hoc
       identifiers in this column to indicate a group of graphemes, which can
       then be used in the description of the graphemes or the context. The
       identifiers should of course be chosen so that they do not conflate
       with any symbols used in the orthography. Note that such
       classes only refer to the graphemes, not to the context. 
	\item \textsc{Columns describing transliterations for each graphemes can be
       added and named at will}. Often more than a single possible
       transliteration will be of interest. Any software application using these
       profiles should prompt the user to name any of these columns to select a
       specific transliteration. 
	\item \textsc{Any other columns can be added freely, but will mostly be ignored
       by any software application using the profiles}. As orthography profiles
       are also intended to be read and interpreted by humans, it is often
       highly useful to add extra information on the graphemes in further
       columns, such as Unicode code points, Unicode names, frequency of
       occurrence, examples of occurrence, explanation of contextual
       restrictions, or comments. 
 \end{enumerate}

@@

\begin{enumerate}
	\def\labelenumi{B\arabic{enumi}.} 
	\item \textsc{Each line of a profile will be interpreted according to the content type of the column as specified in the profile metadata}. Content types include literal and regular expression.
	\item \textsc{The \textsc{class} column will be used to produce explicit
       \textsc{or} chains of regular expressions}, which will then be inserted
       in the \texttt{Grapheme}, \texttt{Left} and \texttt{Right} columns at
       the position indicated by the class-identifiers. For example, a class
       called \texttt{V} as a context specification might be replaced by a regular
       expression like:
       \texttt{(a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au)}.
       Only the graphemes themselves are included here, not any contexts
       specified for the elements of the class. Note that in some cases the 
       ordering inside this regular expression might be crucial, therefore it must be 
	   specified as such in the metadata file.
	\item \textsc{The \textsc{left} and \textsc{right} contexts will be included
       into the regular expressions by using lookbehind and lookahead}.
       Basically, the actual regular expression syntax of lookbehind and
       lookahead is simply hidden to the users by allowing them to only specify
       the contexts themselves. Internally, the contexts in the columns
       \texttt{Left} and \texttt{Right} are combined with the column
       \texttt{Grapheme} to form a complex regular expression like:\ 
       \texttt{(?\textless{}=Left)Grapheme(?=Right)}. 
	\item \textsc{The regular expressions will be applied in the order as specified
       in the profile, from top to bottom.} A software implementation can offer
       help in figuring out the optimal ordering of the regular expressions, but
       then it should be made explicit in the orthography profile because regular 
	   expressions are executed in order from top to bottom.
\end{enumerate}

@@

\begin{enumerate}
	\def\labelenumi{B\arabic{enumi}.} \setcounter{enumi}{4} 
	\item \textsc{All graphemes are matched in the text before they are tokenized
       or transliterated}. In this way, there is no necessity for the user to
       consider \textit{feeding} and \textit{bleeding} situations, in which the application of
       a rule either changes the text so another rule suddenly applies (feeding)
       or prevents another rule from applying (bleeding). 
	\item \textsc{The matching of the graphemes can occur either globally or
       linearly.} From a computer science perspective, the most natural way to
       match graphemes from a profile in some text is by walking linearly
       through the text-string from left to right, and at each position going
       through all graphemes in the profile to see which one matches, then go to
       the position at the end of the matched grapheme and start over. This is
       basically how a finite state transducer works, which is a
       well-established technique in computer science. However, from a
       linguistic point of view, our experience is that most linguists find it
       more natural to think from a global perspective. In this approach, the
       first grapheme in the profile is matched everywhere in the text-string
       first, before moving to the next grapheme in the profile. Theoretically,
       these approaches will lead to different results, though in practice of
       actual natural language orthographies they almost always lead to the same
       result. Still, we suggest that any software application using orthography
       profiles should offer both approaches (i.e.\ \textsc{global} or
       \textsc{linear}) to the user. The approach used should be documented in
       the metadata as \textsc{tokenization method}. 
	\item \textsc{The matching of the graphemes can occur either in NFC or NFD.} By
       default, both the profile and the text-string to be tokenized should be
       treated as NFC (see Section~\ref{pitfall-canonical-equivalence}). However, in some use-cases it turns out to
       be practical to treat both text and profile as NFD. This typically
       happens when very many different combinations of diacritics occur in the
       data. An NFD profile can then be used to first check which individual
       diacritics are used, before turning to the more cumbersome inspection of
       all combinations. We suggest that any software application using
       orthography profiles should offer both approaches (i.e.\ \textsc{NFC} or
       \textsc{NFD}) to the user. The approach used should be documented in the
       metadata as \textsc{unicode normalization}. 
	\item \textsc{The text-string is always returned in tokenized form} by
       separating the matched graphemes by a user-specified symbols-string. Any
       transliteration will be returned on top of the tokenization. 
	\item \textsc{Leftover characters}, i.e.~\textsc{characters that are not
       matched by the profile, should be reported to the user as errors.}
       Typically, the unmatched characters are replaced in the tokenization by a
       user-specified symbol-string.       
 \end{enumerate}

@@

\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item \textsc{should offer user-options} to specify:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} 
		\item \textsc{the name of the column to be used for transliteration} (if any). 
		\item \textsc{the symbol-string to be inserted between graphemes.} Optionally,
        a warning might be given if the chosen string includes characters from
        the orthography itself. 
		\item \textsc{the symbol-string to be inserted for unmatched strings} in the
        tokenized and transliterated output. 
		\item \textsc{the tokenization method}, i.e.~whether the tokenization should
        proceed as \textsc{global} or \textsc{linear} (see B6 above). 
		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile
        should use \textsc{NFC} or \textsc{NFD}. 
    \end{enumerate}
	\item \textsc{might offer user-options}:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{5} 
		\item \textsc{to assist in the ordering of the graphemes.} In our experience 
		working with idiosyncratic transcriptions and orthographies from low-resource languages, it
        is helpful to identify multi-sequence graphemes before single graphemes, and to
        identify graphemes with context before graphemes without context. Further,
        frequently relevant rules might be applied after rarely relevant rules
        (though frequency is difficult to establish in practice, as it depends
        on the available data). Also, if this all fails to give any decisive
        ordering between rules, it seems useful to offer linguists the option to
        reverse the ordering from any manual specified ordering, because
        linguists tend to write the more general rule first, before turning to
        exceptions or special cases. 
		\item \textsc{to assist in dealing with upper and lower case characters.} It
        seems practical to offer some basic case matching, so characters like
        <a> and <A> are treated equally. This will be useful in many concrete
        cases (such as search or collation), although the user should be warned that case matching does not
        function universally in the same way across orthographies.\footnote{For example 
		compare the different first-letter capitalization practices of the digraphs 
		<ǋ> and <Ĳ> (single-character ligatures in the Unicode Standard) 
		in the Latin-based scripts of Southern-Slavic languages and 
		Dutch, respectively.} Ideally,
        users should prepare orthography profiles with all lowercase and
        uppercase variants explicitly mentioned, so by default no case matching
        should be performed. 
		\item \textsc{to treat the profile literally}, i.e.~to not interpret regular
        expression metacharacters. Matching graphemes literally often leads to
        significant speed increase, and ensures that users do not have to worry
        about escaping metacharacters. However, in our experience all actually
        interesting use-cases of orthography profiles include some contexts,
        which automatically prevents any literal interpretation.
    \end{enumerate}
	\item \textsc{should return the following information} to the user:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{8} 
		\item \textsc{the original text-strings} to be processed in the specified
        Unicode normalization, i.e.~in either NFC or NFD as specified by the
        user. 
		\item \textsc{the tokenized strings}, with additionally any transliterated
        strings, if transliteration is requested. 
		\item \textsc{a survey of all errors encountered}, ideally both (i) in which
        text-strings any errors occurred and (ii) which characters in the
        text-strings lead to errors. 
		\item \textsc{a reordered profile}, when any automatic reordering is offered. 
	\end{enumerate}
\end{enumerate}

>>>>

The Unicode Standard offers a very detailed technical approach 
for characterizing writing systems computationally. As such, it is 
sometimes too complex for the day-to-day practice of many linguists, 
as exemplified by the need to understand the common pitfalls that we discussed in Chapters 
\ref{unicode-pitfalls} \& \ref{ipa-meets-unicode}. Therefore, in this section we propose some simple 
guidelines for linguists working in multilingual environments.

Our aims for adopting a Unicode-based
solution are: (i) to improve the consistency of the encoding of sources, (ii)
to transparently document knowledge about the writing system (including
transliteration), and (iii) to do all of that in a way that is easy and quick to
manage for many different sources with many different writing systems. The
central concept in our proposal is the \textsc{orthography profile}, a simple
delimited text file, that characterizes and documents a writing system.
We also offer basic implementations in Python and R to assist with the
production of such files, and to apply orthography profiles for consistency
testing, grapheme tokenization and transliteration. Not only can orthography
profiles be helpful in the daily practice of linguistics, they also succinctly
document the orthographic details of a specific source, and, as such, might
fruitfully be published alongside sources (e.g.~in digital archives). Also, in
high-level linguistic analyses in which the graphemic detail is of central
importance (e.g.~phonotactic or comparative-historical studies), orthography
profiles can transparently document the decisions that have been taken in the
interpretation of the orthography in the sources used.

Given these goals, Unicode Locales (see
Chapter~\ref{the-unicode-approach}) might seem like the ideal orthography
profiles. However, there are various practical obstacles preventing the use of
Unicode Locales in the daily linguistic practice, namely: (i) the
XML structure\footnote{\url{http://unicode.org/reports/tr35/}} is too verbose to easily and quickly produce or correct manually,
(ii) Unicode Locales are designed for a wide scope of information (like date
formats or names of weekdays) most of which is not applicable for documenting
writing systems, and (iii) most crucially, even if someone made the effort to
produce a technically correct Unicode Locale for a specific source at hand,
then it is well-nigh impossible to deploy the description. This is because a locale
description has to be submitted to and accepted by the Unicode Common Locale
Data Repository. The repository is (rightly so) not interested in descriptions
that only apply to a limited set of sources (e.g.~descriptions for only a single
dictionary).

@@

The crucial difference between our current proposal and traditional
computational approaches to transliteration is the strict separation between
tokenization and transliteration. Most computational approaches to
transliteration are based on finite-state transducers (including the
transliteration as described in the Unicode Locale Data Markup
Language).\footnote{\url{http://www.unicode.org/reports/tr35/}} Finite-state 
transducers attempt to describe the mapping from input to output string directly 
as a set of rewrite rules. Although such systems are computationally well 
understood, we feel that they are not well-suited for day-to-day linguistic 
practice. First, by forcing a first step of grapheme tokenization, our system 
tries to keep close to the logic of the writing system. Second, by separating 
tokenization from transliteration there is no problem with `feeding' and 
`bleeding' of rules, common with transducers (cf.~Section~\ref{r-implementation}).

@@

In practice, we foresee a workflow in which orthography profiles are iteratively
refined, while at the same time inconsistencies and errors in the data to be
tokenized are corrected. In some more complex use cases there might even be a
need for multiple different orthography profiles to be applied in sequence (see
Sections~\ref{python-implementations} \& \ref{r-implementation} on various exemplary use cases). The result of any such
workflow will normally be a cleaned dataset and an explicit description of the
orthographic structure in the form of an orthography profile. Subsequently, the
orthography profiles can be easily distributed in scholarly channels alongside
the cleaned data, for example in supplementary material added to journal papers
or in electronic archives.

@@

\begin{enumerate}
	\def\labelenumi{A\arabic{enumi}.} 
	\item \textsc{A profile is a unicode utf-\scalebox{0.8}{8} encoded text file} 
	   % normalized following NFC (or NFD if specified in the metadata)
       that includes information pertinent to the orthography.\footnote{See 
	   Section~\ref{pitfall-file-formats} in which we suggest to use NFC, 
	   no-BOM and LF line breaks because of the pitfalls they avoid. A keen reviewer notes, however, that specifying 
	   a convention for line endings and BOM is overly strict because most 
	   computing environments (now) transparently handle both alternatives. 
	   For example, using Python a file can be decoded using the encoding 
	   ``utf-8-sig'', which strips away the BOM (if present) and reads 
	   an input full in text mode, so that both line feed variants ``LF'' and 
	   ``CRLF'' will be stripped.}

@@

\begin{enumerate}
	\def\labelenumi{A\arabic{enumi}.} \setcounter{enumi}{4} 
	\item \textsc{A minimal profile consists of a single column} with a header
       called \texttt{Grapheme}, listing each of the different graphemes in a
       separate line. The name of this column is crucial for automatic 
       processing.
	\item \textsc{Optional columns can be used to specify the left and right
       context of the grapheme}, to be designated with the headers \texttt{Left}
       and \texttt{Right} respectively. The same grapheme can occur multiple
       times with different contextual specifications, for example to
       distinguish different pronunciations depending on the context. 
	\item \textsc{The columns \texttt{Grapheme}, \texttt{Left} and \texttt{Right}
       can use regular expression \\ metacharacters.} If regular expressions are
       used, then they must be specified in the metadata file as such, 
	   and all literal usage of the special symbols, like full stops <.>
       or dollar signs <\$> (so-called \textsc{metacharacters}) have to be
       explicitly escaped by adding a backslash before them (i.e.~use
       <\textbackslash.> or <\textbackslash\$>). Note that any specification of
       context automatically expects regular expressions, so it is 
       better to always escape all regular expression metacharacters when used
       literally in the orthography. The following symbols will need to be
       preceded by a backslash: {[} {]} ( ) \{ \} | ~+ * . - ! ? \^{} \$ and the
       backslash \textbackslash~itself. 
	\item \textsc{An optional column can be used to specify classes of graphemes},
       to be identified by the header \texttt{Class}. For example, this column
       can be used to define a class of vowels. Users can simply add ad-hoc
       identifiers in this column to indicate a group of graphemes, which can
       then be used in the description of the graphemes or the context. The
       identifiers should of course be chosen so that they do not conflate
       with any symbols used in the orthography. Note that such
       classes only refer to the graphemes, not to the context. 
	\item \textsc{Columns describing transliterations for each graphemes can be
       added and named at will}. Often more than a single possible
       transliteration will be of interest. Any software application using these
       profiles should prompt the user to name any of these columns to select a
       specific transliteration. 
	\item \textsc{Any other columns can be added freely, but will be typically ignored
       by any software application using the profiles}. As orthography profiles
       are also intended to be read and interpreted by humans, it is often
       very useful to add extra information about the graphemes in further
       columns, such as Unicode code points, Unicode names, frequency of
       occurrence, examples of occurrence, explanation of contextual
       restrictions, or comments. 
 \end{enumerate}

@@

\begin{enumerate}
	\def\labelenumi{B\arabic{enumi}.} 
	\item \textsc{Each line of a profile will be interpreted according to the content type of the column as specified in the profile metadata}. Content types include literal and regular expression.
	\item \textsc{The \textsc{class} column will be used to produce explicit
       \textsc{or} chains of regular expressions}, which will then be inserted
       in the \texttt{Grapheme}, \texttt{Left} and \texttt{Right} columns at
       the position indicated by the class-identifiers. For example, a class
       called \texttt{V} as a context specification might be replaced by a regular
       expression like:
       \texttt{(au\textbar{}ei\textbar{}a\textbar{}e\textbar{}i\textbar{}o\textbar{}u}).
       Only the graphemes themselves are included here, not any contexts
       specified for the elements of the class. Note that the 
       ordering inside this regular expression is crucial (e.g.\ regular expressions are greedy, so
	   longest matches should be placed before matching substrings).
	\item \textsc{The \textsc{left} and \textsc{right} contexts will be included
       into the regular expressions by using lookbehind and lookahead}.
       Basically, the actual regular expression syntax of lookbehind and
       lookahead is simply hidden to the users by allowing them to only specify
       the contexts themselves. Internally, the contexts in the columns
       \texttt{Left} and \texttt{Right} are combined with the column
       \texttt{Grapheme} to form a complex regular expression like:\\ 
       \texttt{(?\textless{}=Left)Grapheme(?=Right)}. 
	\item \textsc{The regular expressions will be applied in the order as specified
       in the profile, from top to bottom.} A software implementation can offer
       help in figuring out the optimal ordering of the regular expressions, but
       then it should be made explicit in the orthography profile because regular 
	   expressions are executed in order from top to bottom.
\end{enumerate}

@@

\begin{enumerate}
	\def\labelenumi{B\arabic{enumi}.} \setcounter{enumi}{4} 
	\item \textsc{All graphemes are matched in the text before they are tokenized
       or transliterated}. In this way, there is no necessity for the user to
       consider \textit{feeding} and \textit{bleeding} situations, in which the application of
       a rule either changes the text so another rule suddenly applies (feeding)
       or prevents another rule from applying (bleeding). 
	\item \textsc{The matching of the graphemes can occur either globally or
       linearly.} From a computer science perspective, the most natural way to
       match graphemes from a profile in some text is by walking linearly
       through the text-string from left to right, and at each position going
       through all graphemes in the profile to see which one matches, then go to
       the position at the end of the matched grapheme and start over. This is
       basically how a finite state transducer works, which is a
       well-established technique in computer science. However, from a
       linguistic point of view, our experience is that most linguists find it
       more natural to think from a global perspective. In this approach, the
       first grapheme in the profile is matched everywhere in the text-string
       first, before moving to the next grapheme in the profile. Theoretically,
       these approaches will lead to different results, though in practice of
       actual natural language orthographies they almost always lead to the same
       result. Still, we suggest that any software application using orthography
       profiles should offer both approaches (i.e.\ \textsc{global} or
       \textsc{linear}) to the user. The approach used should be documented in
       the metadata as \textsc{tokenization method}. 
	\item \textsc{The matching of the graphemes can occur either in nfc or nfd.} 
	   The Unicode Standard states that software is free to compose or decompose the character stream from one 
	   representation to another. However, Unicode conformant software must treat canonically equivalent sequences in 
	   NFC and NFD as the same. It is up to the orthography profile creator how they choose to encode their profile. 
	   Several sources suggest to use NFC when possible for text encoding,\footnote{\url{http://www.win.tue.nl/~aeb/linux/uc/nfc_vs_nfd.html}} 
	   including SIL International with regard to data archiving.\footnote{\url{http://scripts.sil.org/cms/scripts/page.php?item_id=NFC_vs_NFD}} 
	   In our experience, in some use cases it turns out to
       be practical to treat both text and profile as NFD. This typically
       happens when many different combinations of diacritics occur in the
       data. An NFD profile can then be used to first check which individual
       diacritics are used, before turning to the more cumbersome inspection of
       all combinations. We suggest that any software application using
       orthography profiles should offer both approaches (i.e.\ \textsc{NFC} or
       \textsc{NFD}) to the user. The approach used can be documented in the
       metadata as \textsc{unicode normalization}. 
	\item \textsc{The text-string is always returned in tokenized form} by
       separating the matched graphemes by a user-specified symbols-string. Any
       transliteration will be returned on top of the tokenization. 
	\item \textsc{Leftover characters}, i.e.~\textsc{characters that are not
       matched by the profile, should be reported to the user as errors.}
       Typically, the unmatched characters are replaced in the tokenization by a
       user-specified symbol-string.       
 \end{enumerate}

@@

\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item \textsc{should offer user-options} to specify:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} 
		\item \textsc{the name of the column to be used for transliteration} (if any). 
		\item \textsc{the symbol-string to be inserted between graphemes.} Optionally,
        a warning might be given if the chosen string includes characters from
        the orthography itself. 
		\item \textsc{the symbol-string to be inserted for unmatched strings} in the
        tokenized and transliterated output. 
		\item \textsc{the tokenization method}, i.e.~whether the tokenization should
        proceed as \textsc{global} or \textsc{linear} (see B6 above). 
		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile
        should use \textsc{NFC} or \textsc{NFD}. 
    \end{enumerate}
	\item \textsc{might offer user-options}:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{5} 
		\item \textsc{to assist in the ordering of the graphemes.} In our experience 
		working with idiosyncratic transcriptions and orthographies from low-resource languages, it
        is helpful to identify multi-sequence graphemes before single graphemes, and to
        identify graphemes with context before graphemes without context. Further,
        frequently relevant rules might be applied after rarely relevant rules
        (though frequency is difficult to establish in practice, as it depends
        on the available data). Also, if this all fails to give any decisive
        ordering between rules, it seems useful to offer linguists the option to
        reverse the ordering from any manual specified ordering, because
        linguists tend to write the more general rule first, before turning to
        exceptions or special cases. 
		\item \textsc{to assist in dealing with upper and lower case characters.} It
        seems practical to offer some basic case matching, so characters like
        <a> and <A> are treated equally. This will be useful in many concrete
        cases (such as search or collation), although the user should be warned that case matching does not
        function universally in the same way across orthographies.\footnote{For example 
		compare the different first-letter capitalization practices of the digraphs 
		<ǋ> and <Ĳ> (single-character ligatures in the Unicode Standard) 
		in the Latin-based scripts of Southern-Slavic languages and 
		Dutch, respectively.} Ideally,
        users should prepare orthography profiles with all lowercase and
        uppercase variants explicitly mentioned, so by default no case matching
        should be performed. 
		\item \textsc{to treat the profile literally}, i.e.~to not interpret regular
        expression metacharacters. Matching graphemes literally often leads to
        significant speed increase, and ensures that users do not have to worry
        about escaping metacharacters. However, in our experience all actually
        interesting use cases of orthography profiles include some contexts,
        which automatically prevents any literal interpretation.
    \end{enumerate}
	\item \textsc{should return the following information} to the user:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{8} 
		\item \textsc{the original text-strings} to be processed in the specified
        Unicode normalization, i.e.~in either NFC or NFD as specified by the
        user. 
		\item \textsc{the tokenized strings}, with additionally any transliterated
        strings, if transliteration is requested. 
		\item \textsc{a survey of all errors encountered}, ideally both (i) in which
        text-strings any errors occurred and (ii) which characters in the
        text-strings lead to errors. 
		\item \textsc{a reordered profile}, when any automatic reordering is offered. 
	\end{enumerate}
\end{enumerate}

### 2018-06-17 book/chapters/pitfalls.tex
closes #30

In this context, it is important to realize that the Unicode Standard was not
developed to solve linguistic problems per se, but to offer a consistent
computational environment for written language. In those cases in which the
Unicode Standard behaves differently than expected, we think it is important not
to dismiss Unicode as wrong or deficient, because our
experience is that in almost all cases the behavior of the Unicode Standard has
been particularly well thought through. The Unicode Consortium has a 
wide-ranging view of matters and often examines important practical use-cases
that are not normally considered from a linguistic point of view. Our general
guideline for dealing with the Unicode Standard is to accept it as it is, and
not to tilt at windmills. Alternatively, of course, it is possible to actively
engage in the development of the standard itself, an effort that is highly
appreciated by the Unicode Consortium.

>>>>

In this context, it is important to realize that the Unicode Standard was not
developed to solve linguistic problems per se, but to offer a consistent
computational environment for written language. In those cases in which the
Unicode Standard behaves differently than expected, we think it is important not
to dismiss Unicode as wrong or deficient, because our
experience is that in almost all cases the behavior of the Unicode Standard has
been particularly well thought through. The Unicode Consortium has a 
wide-ranging view of matters and often examines important practical use cases
that are not normally considered from a linguistic point of view. Our general
guideline for dealing with the Unicode Standard is to accept it as it is, and
not to tilt at windmills. Alternatively, of course, it is possible to actively
engage in the development of the standard itself, an effort that is highly
appreciated by the Unicode Consortium.
